---
title: "Cyclistic Exercise Full Year Analysis"
author: "Yeasin"
date: "7/6/2021"
output: html_document
---
This analysis is for case study 1 from the Google Data Analytics Certificate (Cyclistic).  It’s originally based on the case study "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman (found here: https://artscience.blog/home/divvy-dataviz-case-study). We will be using the Divvy dataset for the case study. The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: “In what ways do members and casual riders use Divvy bikes differently?
[click here to see in Rpub](https://rpubs.com/yeasinopu/Cyclistic)

## Data processing
#### Load Required Package
For downloading data [click Here](https://divvy-tripdata.s3.amazonaws.com/index.html)
```{r library ,message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
getwd()
```
#### Read data
```{r message=FALSE, warning=TRUE, cache=TRUE}
q2_2019 <- read_csv("data/Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("data/Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("data/Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("data/Divvy_Trips_2020_Q1.csv")

```

#### Wrangle data 
Compare column names each of the files. While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file

```{r Rename Columns, cache=TRUE}
q4_2019 <- rename(q4_2019,
      ride_id = trip_id
      ,rideable_type = bikeid 
      ,started_at = start_time  
      ,ended_at = end_time  
      ,start_station_name = from_station_name 
      ,start_station_id = from_station_id 
      ,end_station_name = to_station_name 
      ,end_station_id = to_station_id 
      ,member_casual = usertype
)
q3_2019 <- rename(q3_2019
        ,ride_id = trip_id
        ,rideable_type = bikeid 
        ,started_at = start_time  
        ,ended_at = end_time  
        ,start_station_name = from_station_name 
        ,start_station_id = from_station_id 
        ,end_station_name = to_station_name 
        ,end_station_id = to_station_id 
        ,member_casual = usertype
)

q2_2019 <- rename(q2_2019
    ,ride_id = "01 - Rental Details Rental ID"
    ,rideable_type = "01 - Rental Details Bike ID" 
    ,started_at = "01 - Rental Details Local Start Time"  
    ,ended_at = "01 - Rental Details Local End Time"  
    ,start_station_name = "03 - Rental Start Station Name" 
    ,start_station_id = "03 - Rental Start Station ID"
    ,end_station_name = "02 - Rental End Station Name" 
    ,end_station_id = "02 - Rental End Station ID"
    ,member_casual = "User Type"
)

```
#### Convert ride_id and rideable_type to character so that they can stack correctly

```{r Mutate columns and binding}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id) 
                   ,rideable_type = as.character(rideable_type))
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
rm(q2_2019)
rm(q3_2019)
rm(q4_2019)
rm(q1_2020)
all_trips <-all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration")) %>%
  as.data.frame()

```
## CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
```{r}
summary(all_trips)
glimpse(all_trips)
skim_without_charts(all_trips)
table(all_trips$member_casual)
```

In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"
```{r}
all_trips <- all_trips %>% mutate(member_casual = recode(member_casual,
                                            "Subscriber"= "member",
                                            "Customer" = "casual"))
table(all_trips$member_casual)
```

Add columns that list the date, month, day, and year of each ride. This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level
```{r date}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(all_trips$date, "%m")
all_trips$day <- format(all_trips$date, "%d")
all_trips$year <- format(all_trips$date, "%Y")
all_trips$day_of_week <- format(all_trips$date, "%A")
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))

```

#### Remove "bad" data
The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative.We will create a new version of the dataframe (v2) since data is being removed
```{r}
select(all_trips,ride_length, ended_at,started_at, start_station_name) %>%
  filter(ride_length < 0) %>% head(30)
all_trips_v2<- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length < 0),]
```
#### CONDUCT DESCRIPTIVE ANALYSIS
Compare members and casual users:
```{r}
rm(all_trips)
summary(all_trips_v2$ride_length)
all_trips_v2 %>% group_by(member_casual) %>%
  summarise(mean_ride_length = mean(ride_length), median_ride_length = median(ride_length), max_ride_length = max(ride_length), min_ride_length = min(ride_length))
```

See the average ride time by each day for members vs casual users
```{r}
avg_trip_summary <- all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>%
  summarise(number_of_ride = n(), avg_ride_length = mean(ride_length), .groups = "drop")
avg_trip_summary
```
Let's visualize the number of rides by rider type:
```{r}
ggplot(data = avg_trip_summary, aes(x=day_of_week, y= number_of_ride, fill = member_casual)) + geom_col(position = "dodge")
```

